---
title: "EasternCanadaDataPrep Manual"
subtitle: "Spatial data preparation for Eastern Canada"
date: "Last updated: 2026-02-03"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    theme: sandstone
    number_sections: false
---

```{r setup, include=FALSE}
library(SpaDES.core)
library(SpaDES.project)
library(terra)
library(sf)

knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Executive summary

EasternCanadaDataPrep is a SpaDES module that prepares and harmonizes
core spatial inputs for forest modelling applications in Eastern Canada.

The module defines a spatial planning framework, harmonizes spatial
layers to a common grid, and identifies areas that are legally available
for harvesting, prior to applying ecological or management rules.

This module is intentionally limited to spatial harmonization and legal
constraints only. It does not interpret forest condition, stand dynamics,
or management policies.

---

## Purpose

This module answers a single question:

> Where are spatial planning units defined, and which areas are legally
> available for harvesting before ecological or management rules are applied?

To answer this question, the module:

- Defines a consistent Eastern Canada study area  
- Harmonizes all spatial inputs to a common projection  
- Prepares administrative boundaries (FMUs)  
- Applies legal exclusions (protected and conserved areas)  
- Constructs a coarse-resolution PlanningGrid_250m  
- Aligns LandCover and StandAge rasters to the planning grid  

---

## What this module does

✔ Defines a study area for Eastern Canada  
✔ Loads and harmonizes Forest Management Units (FMUs)  
✔ Loads and filters protected and conserved areas (CPCAD)  
✔ Constructs a PlanningGrid_250m (default: 250 m)  
✔ Aligns land-cover raster (if supplied)  
✔ Loads or downloads SCANFI 2020 stand age  
✔ Creates a LegalHarvestMask_250m  

---

## What this module does not do

This module intentionally avoids ecological or management interpretation.

It does not:

- Classify forest vs. non-forest  
- Estimate biomass, productivity, or growth  
- Apply riparian buffering  
- Apply silvicultural or management rules  
- Allocate or schedule harvest  

All such decisions are deferred to downstream modules.

---

## Spatial reference system

All spatial inputs and outputs are harmonized to:

Canada Albers Equal Area Conic  
ESRI:102001

---

## Input objects

| Object | Class | Description |
|--------|--------|-------------|
| `studyArea` | `sf` / `SpatVector` | Polygon defining the modelling extent |
| `FMU` | `sf` / `SpatVector` | Forest Management Unit boundaries |
| `CPCAD` | `sf` / `SpatVector` | Protected and conserved areas |
| `LandCover` | `SpatRaster` (optional) | Land cover raster supplied by upstream module |
| `standAgeMap` | `SpatRaster` (optional) | SCANFI 2020 stand age raster |

If not provided by the user, required vector inputs are automatically
created or downloaded inside `.inputObjects()`.

---

## Study area definition

If no study area is supplied, the module constructs a default
Eastern Canada study area including:

- Ontario  
- Québec  
- New Brunswick  
- Nova Scotia  
- Prince Edward Island  
- Newfoundland and Labrador  

All layers are cropped, masked, and reprojected to the project CRS.

---

## Forest Management Units (FMUs)

Forest Management Units are:

- Loaded as vector polygons  
- Cropped and masked to the study area  
- Reprojected to ESRI:102001  
- Rasterized to the planning grid  

Rasterization uses a center-of-pixel rule to ensure consistent
assignment of planning cells to management units.

FMUs define the administrative units used by downstream accounting
and harvest-related modules.

---

## Protected and conserved areas (CPCAD)

Protected areas are derived from the
Canadian Protected and Conserved Areas Database (CPCAD).

Processing includes:

- Cropping and masking to the study area  
- Reprojection to ESRI:102001  
- Policy-level filtering only  

Filtered CPCAD areas represent legal exclusions from harvesting.

---

## Land cover — upstream dependency

Land cover is treated as an optional upstream dependency.

If supplied, it must be provided as:

- Object name: `sim$LandCover`  
- Type: `terra::SpatRaster`  

If not supplied, a temporary synthetic raster is created for standalone testing.

Within EasternCanadaDataPrep, land cover is handled spatially only:

- Reprojection to the project CRS  
- Cropping and masking to the study area  
- Aggregation or resampling to 250 m  

The harmonized raster is exposed as:

`sim$LandCover_250m`

---

## Stand age — SCANFI 2020

Stand age is based on:

SCANFI_att_age_S_2020_v1_1.tif

The module uses a three-level fallback logic:

1. If supplied upstream → used directly  
2. If found locally → loaded  
3. Otherwise → automatically downloaded from a project Drive repository  

The raster is then:

- Reprojected if necessary  
- Cropped to the study area  
- Aggregated or resampled to 250 m  

The aligned output is exposed as:

`sim$standAge_250m`

---

## PlanningGrid_250m

The PlanningGrid_250m is a coarse-resolution grid (default: 250 m)
that defines the spatial units used by downstream models.

It is constructed from:

- The study area extent  
- A fixed target resolution  
- The project CRS  

No ecological interpretation is applied at this stage.

---

## LegalHarvestMask_250m

A binary LegalHarvestMask_250m is created where:

- Cells fall inside an FMU  
- Cells are not inside protected or conserved areas  

This mask represents legal availability only.

---

## Output objects

| Output | Type | Description |
|--------|--------|-------------|
| PlanningGrid_250m | SpatRaster | Coarse-resolution planning grid |
| LegalConstraints | list | FMU raster, CPCAD raster, legal harvest mask |
| LandCover_250m | SpatRaster | Harmonized land cover (if supplied) |
| standAge_250m | SpatRaster | Harmonized SCANFI 2020 stand age |

Structure of legal constraints:

```{r, eval=FALSE}
sim$LegalConstraints <- list(
  FMU_Raster_250m        = "<SpatRaster>",
  CPCAD_Raster_250m      = "<SpatRaster>",
  LegalHarvestMask_250m  = "<SpatRaster>"
)
```

---

## Example usage

```{r, eval=FALSE}
rm(list = ls())
gc()

library(SpaDES.core)
library(SpaDES.project)
library(terra)

setPaths(
  cachePath   = "E:/EasternCanadaDataPrep/cache",
  inputPath   = "E:/EasternCanadaDataPrep/inputs",
  outputPath  = "E:/EasternCanadaDataPrep/outputs",
  modulePath  = "E:/EasternCanadaDataPrep/modules",
  scratchPath = "E:/EasternCanadaDataPrep/scratch"
)

getModule(
  modules    = "shirinvark/EasternCanadaDataPrep",
  modulePath = getPaths()$modulePath,
  overwrite  = TRUE
)

rstLCC <- terra::rast(
  "E:/MODULES_TESTS/SCANFI_att_nfiLandCover_CanadaLCCclassCodes_S_2010_v1_1.tif"
)

sim <- simInit(
  times   = list(start = 1, end = 1),
  modules = "EasternCanadaDataPrep",
  objects = list(
    LandCover = rstLCC
  )
)

sim <- spades(sim)

names(sim)
names(sim$LegalConstraints)
```

---

## Relationship to downstream modules

This module must be run before:

- RiparianBuffers  
- LandR forest dynamics modules  
- AAC calculation modules  
- Harvest scheduling modules  

It provides spatial and legal constraints only.

---

## Summary

EasternCanadaDataPrep defines where harvest could legally occur,
not where it should occur.

This separation ensures a clean, modular, and scalable forest modelling architecture.